---
title: "0X_SIR calculation"
author: "Brenna Kelly"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}

library(sf)
library(tmap)
library(tidyr)
library(dplyr)
library(purrr)
library(ggplot2)
library(stringr)
library(SpatialEpi)
# library(data.table)

```

adding cluster assignments to birth data, based on grid id and study week:

```{r}


cluster_week <- read.csv("data/clean_data/clusters_by_gridweek_v2.csv")
full_dat <- read.csv("data/births/clean_birth_data.csv")


# birth-grids with start week
grid <- fread("data/births/gridded_births_morebw.csv")
# sub grid_exp with full_dat

full_dat$start_week <- full_dat$birth_week - full_dat$gestation

births <- full_dat |>
  # filter(start_week > 0) |>
  select(c(momid, grid_id, birth_week, start_week))

names(births) <- c("momid", "grid_id", "week", "start_week")
births <- births[order(births$week), ]

births$birth_week <- births$week

birth_wks <- births %>%
  mutate(val = NA) %>%
  pivot_wider(id_cols = c(momid, grid_id, start_week, birth_week), names_from = week, values_from = val)

names(birth_wks)

names(birth_wks)[5:213] <- paste0("week_", names(birth_wks)[5:213])

birth_wks <- birth_wks %>%
  mutate(across(week_1:week_209, ~ NA))

cluster_week$momid <- NA
cluster_week$start_week <- NA
cluster_week$birth_week <- NA

fill_vars <- names(birth_wks)[5:213]

x <- rbind(birth_wks, cluster_week) %>%
  group_by(grid_id) %>%
  fill(fill_vars, .direction = "downup") %>%
  drop_na()
head(x)

x <- x |>
  filter(start_week > 0)

for(i in 20417:nrow(x)) {
  # make weeks not in pregnancy NA
  end <- x[[i, "birth_week"]]
  start <- x[[i, "start_week"]]
  x[i, ] <- x[i, ] %>%
    mutate_at(-c(1:4, (start + 4):(end + 4)), ~ NA)
}


print(head(x))

write.csv(x, "data/births/birth_traj_jun08.csv", row.names = FALSE)

```

tidying up:

```{r}

x <- fread("data/births/birth_traj_jun08.csv")

# # make clusters one column
x <- unite(x, col = "all_clusters", c(5:213), sep = ", ", na.rm = TRUE)

x <- x[, c("momid", "grid_id", "birth_week", "all_clusters")]

z <- x |>
  separate(all_clusters, into = paste0("week_", c(1:45)), ", ") |>
  mutate_at(c(4:48), ~ ifelse(is.na(.), 0, .)) |>
  mutate_at(c(4:48), ~ as.factor(.))

write.csv(z, "data/births/birth_traj_trunc_jun08.csv", row.names = FALSE)

```

combining exposures and outcomes

```{r}

# birth data
full_dat <- read.csv("data/clean_birth_data.csv")

# trajectories
z <- read.csv("data/birth_traj_trunc_jun08.csv")

# outcome: placental insufficiency
full_dat$hypertension_code <- ifelse(full_dat$hypertension %in% c("C", "E", "H", "P"),
                                "hyp", "no")
full_dat$pl_insuf <- case_when(full_dat$sga10_by_sex > full_dat$birthweightgrams ~ 1, # sga
                           full_dat$hypertension %in% c("H", "E", "P") ~ 1)
full_dat$pl_insuf <- ifelse(is.na(full_dat$pl_insuf), 0, 1)

# outcome: gestational hypertension
full_dat$gestational_hyp <- ifelse(full_dat$hypertension %in% c("H", "E", "P"), 1, 0)

# outcome: gestational diabetes
full_dat$gestational_diab <- ifelse(full_dat$gestationaldiabetes == 1, 1, 0)

# outcome: preterm birth
full_dat$preterm <- ifelse(full_dat$gestation < 34, "preterm", "term")

# outcome: very preterm birth
full_dat$very_preterm <- ifelse(full_dat$gestation < 30, "preterm", "term")

# *negative control* outcome: gender
full_dat$gender_code <- ifelse(full_dat$gender == "M", 1, 0)


# combine outcome and exposures (clusters)
z_y <- merge(z, full_dat[, c("momid", "grid_id", 
                             "pl_insuf", "preterm",
                             "gestational_hyp", "very_preterm", #"sga_code",
                             "gestational_diab", "gender_code")], 
             by = c("momid", "grid_id"))

```

### aggregating number of disease/disease-free by exposure-week

```{r}

names(z_y)[4:48]

dt_res = data.frame()
for (i in 4:48){

  dt_temp = data.frame(table(z_y[, i], z_y$preterm)) # put the outcome here
  #dt_temp$Var1 = names(z_sga)[i]
  #dt_temp$Var2 = names(z_sga)[i + 1]
  dt_temp$week <- paste(names(z_y)[i])

  dt_res = rbind(dt_res, dt_temp)

}

# here we have the number of cases and non-cases for each pregnancy week and cluster
names(dt_res) <- c("cluster", "y", "freq", "week")

# cluster 0 means you weren't pregnant in that week; remove
dt_res <- dt_res |>
  filter(!cluster == 0)

```


### calculating standardized incidence ratios

Outcomes to investigate further:  
- cluster 7 and preterm birth, weeks 0:20    
- cluster 8 and preterm birth, weeks 20:30   
- cluster 10 and preterm birth, weeks 7:13 (this is the one)  
- cluster 11 and preterm birth, weeks 7:13 (lower risk)   

- cluster 11 and placental insufficiency, weeks 10:20 (2% higher risk; CI prob overlaps)  

negative control
- gender doesn't seem to have a relationship with exposure (...good)  


```{r}

# choose a cluster
cluster_of_interest = 10

dt_res$target_cluster <- ifelse(dt_res$cluster == cluster_of_interest, 
                                "target_cluster", "other_cluster")
dt_res_wide <- dt_res |>
  pivot_wider(names_from = c(y, week), 
              values_from = freq)

target_cluster <- data.frame()

# this loops through each week by case and non-case status
for(i in seq(from = 3, to = 92, by = 2)) {
  var_app <- i
  var_sm <- i + 1

  res <- dt_res_wide[, c(2, var_app, var_sm)]

  res$pop <- rowSums(res[2:3])
  names(res) <- c("cluster", "hyp", "no", "pop")

  E <- expected(
    population = res$pop,
    cases = res$hyp,
    n.strata = 1
  )

  res$E <- E#[match(res[, "Var1"], unique(res[, "Var1"]))]
  res$SIR <- res[, "hyp"] / res[, "E"]
  res$week <- str_sub(names(dt_res_wide)[i + 1], start = (nchar(names(dt_res_wide)[i + 1])-1),
                      end = nchar(names(dt_res_wide)[i + 1]))
  res$week <- gsub("_", "", res$week)
  res$week <- as.numeric(res$week)

  target_cluster <- rbind(target_cluster, res)
}

cluster_5_only <- target_cluster |>
  filter(cluster == "target_cluster") |>
  filter(week < 34)

plot(cluster_5_only$week, cluster_5_only$SIR$hyp,
     xlab = "Pregnancy Week",
     ylab = "outcome SIR", main = paste0("Cluster ", cluster_of_interest))
abline(h = 1, lty = 2)

```


### permutation

```{r}

library(rsample)
set.seed(4325)

# testing, just for a single week and cluster
z_y_week_12 <- z_y |>
  dplyr::select(c("week_12", "preterm")) |>
  mutate(cluster_week = ifelse(week_12 == 10, 1, 0)) |>
  mutate(preterm = ifelse(preterm == "preterm", 1, 0))

perm(z_y_week_12$cluster_week, z_y_week_12)

Dnull = replicate(n = 5000, expr = perm(z_y_week_12$cluster_week, z_y_week_12))

Dobs = mean(z_y_week_12[z_y_week_12$cluster_week == "1", "preterm"]) - mean(z_y_week_12[z_y_week_12$cluster_week == "0", "preterm"])

hist(Dnull, col = "grey")
abline(v = Dobs, col = "blue", lwd = 3, lty = 2)

table(Dnull > Dobs) # 2 in 5000
mean(abs(Dnull) >= Dobs) # p = 8e-0.4 for week 12

# writing a function, to scale this up for all weeks
perm = function(x, y) {
  # turn x to a character, easier to deal with
  x = as.character(x)
  # shuffle the x values:
  x_shuff = sample(x)
  # calculate the mean of each group:
  x_bar_add = mean(y[x_shuff == "1", "preterm"])
  x_bar_ctl = mean(y[x_shuff == "0", "preterm"])
  # calculate the difference:
  x_bar_add - x_bar_ctl
  
}

perm(z_y_week_12$week_12_cluster_10, z_y_week_12)

# Note: this just obtains p-values for differences between individual-level outcomes
# the chunk below uses the individual-level permutations to simulate MC
p_res = data.frame()

# for weeks 1:45, use columns 4:48
for (i in 4:48){
  
  pt_week <- z_y |>
    dplyr::select(c(i, "preterm")) |>
    mutate(preterm = ifelse(preterm == "preterm", 1, 0))
  
  names(pt_week) <- c("cluster_week", "preterm")
  
  pt_week <- pt_week |>
    mutate(cluster_week = ifelse(cluster_week == 10, 1, 0))
  
  obs_diff = mean(pt_week[which(pt_week$cluster_week == 1), "preterm"]) -
    mean(pt_week[which(pt_week$cluster_week != 1), "preterm"])
  
  null_diff = replicate(n = 1000, expr = perm(pt_week$cluster_week, pt_week))
  
  hist(null_diff, col = "grey")
  abline(v = obs_diff, col = "blue", lwd = 3, lty = 2)
  
  table(null_diff > obs_diff)
  
  p_val = mean(abs(null_diff) >= obs_diff) # two-tailed p-value
  print(p_val)
  
  p_res <- rbind(p_res,
                 c(paste0("week_", i - 3),
                   p_val))
}

head(p_res)
# again — it's not the SIR, so let's create that permutation
```

```{r}
#########
# SIR monte-carlo
#########

# function to create permuted cluster assignments and calculate the SIR for each week
perm_cluster <- function(y) { # takes data
  
  dt_res_mc = data.frame()
  
  for (i in 4:48){
    
    x = as.character(y[, i])
    # shuffle the x values:
    x_shuff = sample(x)
    
    dt_temp = data.frame(table(x_shuff, y$preterm)) # put the outcome here
    #dt_temp$Var1 = names(z_sga)[i]
    #dt_temp$Var2 = names(z_sga)[i + 1]
    dt_temp$week <- paste(names(y)[i])
    
    dt_res_mc = rbind(dt_res_mc, dt_temp)
    
  }
  
  dt_temp = data.frame(table(x_shuff, y$preterm)) # outcome
  dt_temp$week <- paste(names(y)[i])
  
  # dt_res_mc = rbind(dt_res_mc, dt_temp)
  
  names(dt_res_mc) <- c("cluster_mc", "y", "freq", "week")
  # print(table(dt_res_mc$cluster_mc))
  
  # cluster 0 means you weren't pregnant in that week; remove
  dt_res_mc <- dt_res_mc |>
    filter(!cluster_mc == 0)
  
  # SIR
  dt_res_mc$target_cluster <- ifelse(dt_res_mc$cluster_mc == 10, 
                                  "target_cluster", "other_cluster")
  
  dt_res_wide_mc <- dt_res_mc |>
    tidyr::pivot_wider(id_cols = c(cluster_mc, target_cluster), 
                       names_from = c(y, week), 
                values_from = freq)
  
  # print(head(dt_res_wide_mc))
  
  target_cluster_mc <- data.frame()
  
  for(i in seq(from = 3, to = 92, by = 2)) {
    var_app <- i
    var_sm <- i + 1
    
    res_mc <- dt_res_wide_mc[, c(2, var_app, var_sm)]
    
    res_mc$pop <- rowSums(res_mc[2:3])
    
    names(res_mc) <- c("cluster", "hyp", "no", "pop")
    
    E <- expected(
      population = res_mc$pop,
      cases = res_mc$hyp,
      n.strata = 1
    )
    
    res_mc$E <- E
    res_mc$SIR <- res_mc[, "hyp"] / res_mc[, "E"]
    
    res_mc$week <- str_sub(names(dt_res_wide_mc)[i + 1], 
                           start = (nchar(names(dt_res_wide_mc)[i + 1])-1),
                           end = nchar(names(dt_res_wide_mc)[i + 1]))
    res_mc$week <- gsub("_", "", res_mc$week)
    res_mc$week <- as.numeric(res_mc$week)
    
    target_cluster_mc <- rbind(target_cluster_mc, res_mc)
    
  } 
  
  target_cluster_mc_sir <- target_cluster_mc |>
    filter(cluster == "target_cluster")
  
  target_cluster_mc_sir$SIR <- target_cluster_mc_sir$SIR$hyp
  
  target_cluster_mc_sir[, c("SIR")]
  
}

# this is one MC simulation:
test <- perm_cluster(z_y)

## 1000 MC simularions:
# null_diff <- replicate(n = 1000, expr = perm_cluster(z_y))
# null_diff_cl <- map_df(null_diff, ~as.data.frame(t(.)))

### aside: can we parallelize this^
library(future.apply)
plan(multisession, workers = (detectCores() - 1))

system.time(
  y <- future_replicate(1000, perm_cluster(z_y))
)
# yes we can. takes about 2.5m

# each row is one MC run
# each column is one week
null_diff_cl <- map_df(y, ~as.data.frame(t(.)))

obs_SIR <- cluster_5_only$SIR$hyp[1]

# pivot, for facet plot
null_diff_cl_long <- null_diff_cl |>
  pivot_longer(V1:V45) |>
  rename(week = name) |>
  mutate(week = as.numeric(
    gsub("V", "", week)))

# observed SIR for each week
cluster_5_only$SIR <- cluster_5_only$SIR$hyp

# combine, to plot both
mc_obs <- merge(null_diff_cl_long, cluster_5_only, by = "week")

ggplot(mc_obs, aes(x = value)) +
  geom_histogram() +
  facet_wrap(~ week) +
  geom_vline(data = . |> group_by(week) |> 
               summarise(avg_sir=mean(SIR)), 
             aes(xintercept=avg_sir))

# snag p-values
for(i in 1:length(names(null_diff_cl))) {
  print( paste0("week_", i, ": ",
    mean(abs(null_diff_cl[, i]) >= cluster_5_only[which(cluster_5_only$week == i), "SIR"]))
  )
}



```

Confidence interval calculation

```{r}

##


```



q
- if we used only one component of the mixture (average exposure in cluster relative to average exposure outside the cluster), would we see the same relationship? does the mixture capture what's missing?
    - if there's a cluster mainly made up of one pollutant?





